---
title: 'ESPM 244: Movement Analysis'
author: "Ian Wang"
date: "November 5, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

####################################################################
##                                                                ##
##    ESPM 244: Workshop 10                                       ##
##    Movement Analysis                                           ##
##                                                                ##
####################################################################

Set up the workspace...
```{r}
library(adehabitatLT)
library(adehabitatHR)
library(BBMM)
library(rgdal)
library(raster)
library(move)

setwd()
```

## Part 1: Basic Movement Trajectories ##

Load the data...
```{r}
muleys <-read.csv("muledeer.csv", header = TRUE)
head(muleys)
```

Check for duplicates in the dataset...
```{r}
summary(duplicated(muleys))
```

We first need to convert the date to POSIX format in order to plot movement trajectories...
```{r}
GPSDate <- as.POSIXct(strptime(muleys$GPSFixTime,format="%Y.%m.%d %H:%M:%S"))
muleys$GPSDate <- GPSDate
head(muleys)
```

Measure the time differences between each recorded point...
```{r}
timediff <- diff(muleys$GPSDate)
muleys <- muleys[-1,] # Remove first row
muleys$TimeDiff <-abs(as.numeric(timediff))
dim(muleys)
head(muleys)
```

See how many recorded locations there are for each animal...
```{r}
summary(muleys$id)
plot(muleys$X, muleys$Y, pch='+')
```

Remove outlier locations or known outliers collected too far apart in time...
```{r}
muleys.sub <- subset(muleys, muleys$X > 599000 & muleys$X < 705000 & muleys$Y > 4167000 & muleys$TimeDiff < 14401)
dim(muleys.sub)
head(muleys.sub)
```

Plot the total set of location records...
```{r}
plot(muleys.sub$X, muleys.sub$Y, pch='+')
```

Or plot each animal with a separate color...
```{r}
plot(muleys.sub$X, muleys.sub$Y, pch='+', col=muleys.sub$id)
```

To create movement trajectories, we need to add the data to a spatial points data frame...
```{r}
xy = muleys.sub[c("X","Y")]
xy.sp <- SpatialPoints(xy) # spatial points object
xy.df <- data.frame(xy.sp) # spatial data frame

move.df <- data.frame(ID = muleys.sub[2], TimeDiff = muleys.sub[24], Burst = muleys.sub[23])
coordinates(move.df) <- xy.df
head(move.df)
```

Visualize the movement point locality data (should be the same as above)...
```{r}
plot(move.df)
```

Now create an object of class "ltraj" by animal using the ID field and display the trajectory for the first deer...
```{r}
ltraj <- as.ltraj(coordinates(move.df), move.df$GPSDate, id=move.df$id)
plot(ltraj[1])
head(ltraj[1]) # Describes the trajectory for the first deer
```

Or plot the trajectories for all of them...
```{r}
plot(ltraj)
```

Let's create a histogram of time lag (i.e., interval) and distance between successive locations for each deer. This is a nice way to inspect the time lag between locations as you don't want to include a location if too much time has passed since the previous and it also shows why a trajectory is irregular.
```{r}
hist(ltraj[1], "dt", freq = TRUE)
hist(ltraj[1], "dist", freq = TRUE)

hist(ltraj[2], "dt", freq = TRUE)
hist(ltraj[2], "dist", freq = TRUE)
```

## Part 2: Brownian Bridge Movement Models ##

Create an object with the movement data for deer D15...
```{r}
deerD6 <- subset(muleys.sub, muleys.sub$id == "D6")
head(deerD6)
```

Use brownian.bridge function (BBMM package) to estimate the movements between recorded points...
```{r}
D6.BBMM = brownian.bridge(x=deerD6$X, y=deerD6$Y, time.lag=deerD15$TimeDiff, location.error=34, cell.size=100)
bbmm.summary(D6.BBMM)
```

Plot the areas covered by 50%...95%...99% of movement paths...
```{r}
contours <- bbmm.contour(BBMM, levels=c(seq(50, 90, by=10), 95, 99), locations=deerD6, plot=TRUE)
```

```{r}
xyz <- data.frame(D6.BBMM$x, D6.BBMM$y, D6.BBMM$probability)
bbmm.map <- rasterFromXYZ(xyz)
plot(bbmm.map)
```

## Part 3: Dynamic Brownian Bridge Movement Models ##

While BBMMs incorporate a temporal component and GPS error into estimates, dynamic Brownian Bridge Movement Models (dBBMM) incorporate temporal and behavioral characteristics of movement paths into estimation of home ranges or utilization distributions.

First, create a Move object for an individual deer...
```{r}
deerD6$DT <-as.POSIXct(strptime(deerD6$GPSFixTime, format="%Y.%m.%d %H:%M:%OS")) # Convert time format
#deerD6 <- deerD6[order(deerD6$DT),]
D6.move <- move(x=deerD6$X, y=deerD6$Y, time=deerD6$DT, proj=CRS("+proj=utm +zone=12 +datum=NAD83"), data=deerD6, animal=deerD6$id)
```

Then run dBBMM for the deer to estimate its utilization distribution on the landscape...
```{r}
D6.dbbmm <- brownian.bridge.dyn(object=D6.move, location.error=22, window.size=19, margin=7, dimSize=100, time.step=180)
D6.dbbmm
```

And plot the results...
```{r}
plot(D6.dbbmm)
contour(D6.dbbmm, levels=c(.5,.9,.95,.99), add=TRUE)
```

To determine the area size of the utilization distribution, we can calculate the contour raster from the UD raster...
```{r}
D6.vol <- getVolumeUD(D6.dbbmm)
D6.95zone <- D6.vol <= .95
plot(D6.95zone, main="95% Utilization Distribution")
area95 <- sum(values(D6.95zone))
area95

D6.75zone <- D6.vol <= .75
plot(D6.75zone)
area75 <- sum(values(D6.75zone))
area75
```

### Challenge 1: Subset the recorded locality data to remove records separated by long time lags.  
### Re-run the BBMMs and dBBMMs to see how this changes the resulting utilization distributions.

### Challenge 2: Examine overlap in the home ranges of different deer.  Construct BBMMs or dBBMMs 
### for all (or at least some) individuals.  Then calculate the overlap in ranges or the cumulative
### utilization of different cells on the raster.